---
title: "2024-04_ev_dilution_counts"
output: github_document
---

# EV dilution - data preprocessing
```{r setup}
# install.packages("Seurat")
# install.packages("remotes")
# install.packages("BiocManager")
# BiocManager::install(version = '3.18')
# BiocManager::install("glmGamPoi")
# remotes::install_github("stephenturner/annotables")
# install.packages("glmGamPoi")
# BiocManager::install("DESeq2")
# # BiocManager::install("MAST") # <- Doesn't work
# remotes::install_github("RGLab/MAST")
# # install.packages("data.filt")
# remotes::install_github("sonejilab/cellexalvrR") # <- Something Yang wanted to try
# remotes::install_github("mojaveazure/seurat-disk")

library(Seurat)
# library(dplyr)
library(data.table)
library(Matrix)
library(ggplot2)
library(scales) # to better customize Seurat's plots
library(ggpubr)
library(ggrepel)
library(patchwork)
library(cowplot)
library(ggbeeswarm) # for better positioning of arrows on log2 change plots - position_jitter doesn't allow to mark points
library(future)
library(annotables) # for turning Ensembl ID to symbol
library(sctransform) # for normalization  
library(glmGamPoi) # for SCTransform
# library(svglite) # for vectorized, lightweight plotting
library(systemfonts) # to set the font for svg outputs
# library(DESeq2)
library(MAST)
# library(cellexalvrR) <- Something Yang wanted to try
library(SeuratDisk)

"%notin%" <- Negate("%in%")
"%notlike%" <- Negate("%like%")


# set the theme for plotting (Seurat uses ggplot's themes)
theme_set(new = theme_classic())
theme_update(
  axis.text.x = element_text(vjust = 0.5),
  strip.background = element_rect(fill = '#FFFFFF'),
  plot.title = element_text(hjust = 0.5, size = 25),
  plot.subtitle = element_text(size = 20, hjust = .5),
  axis.title = element_text(size = 23),
  axis.text = element_text(size = 20),
  legend.text = element_text(size = 18),
  legend.key.size = unit(2, 'line'),
  legend.title = element_text(size = 20, hjust = .5, vjust = .5)
  # text = element_text(family= "mono")
)

# That's not necessary (rmarkdown sets its directory as the one the .Rmd file is in.)
wd <- "/home/radgro/projects/2023-12_dilution_ev_seq/analysis/2024-04_ev_dilution_counts/"
knitr::opts_knit$set(root.dir = wd)

fonts <- list(
  mono = "Consolas",
  sans = "Consolas"
)

# set svglite as a default for all the plots
# knitr::opts_chunk$set(knitr.chunk.dev = 'svglite')
# knitr::opts_chunk$set(dev = 'svglite', system_fonts = fonts)
knitr::opts_chunk$set(dev = 'svglite', dev.args = list(system_fonts = fonts),
                      cache.path = "ev_dilution_counts/gfm", cache = F,
                      cache.lazy = FALSE) # cache of a github_document doesn't work if the path to the gfm folder is not provided!!!

# knitr::opts_chunk$set(cache.extra = 1) # RESETS CACHE

# plan("multicore", workers = 8) # Not allowed on the server
# plan()

```


```{r load the data}
dil_full <- readRDS("../mapping/zUMIs_output/zUMIs_output/expression/2024-04_dilution_ev_seq.dgecounts.rds")
dil <- CreateSeuratObject(dil_full$umicount$inex$all)
```


```{r}
VlnPlot(dil, features = c("nCount_RNA", "nFeature_RNA"))
```

## Decode the sample barcodes and insert metadata
```{r}
barcodes <- fread("/disk2/radgro/projects/2023-12_dilution_ev_seq/data/reads_for_zUMIs.samples.txt")

barcodes[, orig.ident := gsub("[0-9].*", "", r1)]

barcodes[, dilution := gsub("[A-Z][a-z]?", "", r1)
         ][, dilution := gsub("_.*", "", dilution)
           ][, dilution := gsub("100k", "", dilution)]

barcodes[, replicate := gsub("_S[0-9].*", "", r1)
         ][, replicate := gsub(".*_", "", replicate)]

barcodes <- barcodes[, .(orig.ident, dilution, replicate, BC)]
```

```{r order the barcodes}
dil.ident <- data.table("BC" = colnames(dil))
bc.order <- barcodes[dil.ident, on = "BC"] # order the metadata table 

table(bc.order$BC == dil.ident) # are they in the same order?
```

```{r insert the metadata to the Seurat object}
dil@meta.data$orig.ident <- bc.order[, orig.ident]
dil@meta.data$dilution <- bc.order[, dilution]
dil@meta.data$replicate <- bc.order[, replicate]

Idents(dil) <- bc.order[, orig.ident]
```


```{r, fig.width=10, fig.height=10}
p1 <-VlnPlot(dil, features = c("nCount_RNA", "nFeature_RNA"), group.by = "dilution")
p2 <-VlnPlot(dil, features = c("nCount_RNA", "nFeature_RNA"), group.by = "dilution", log = T)
p3 <-VlnPlot(dil, features = c("nCount_RNA", "nFeature_RNA"), group.by = "orig.ident", log = T)

ggarrange(p1, p2, p3, ncol = 1)
  
```


```{r}
meta <- dil@meta.data
setDT(meta)

meta[, mean(nFeature_RNA), by = dilution]
```

```{r, fig.width = 10, fig.height=7.5}
ggplot(meta) +
  geom_jitter(aes(x = orig.ident, y = nFeature_RNA, size = dilution, color = dilution), width = .2) +
  scale_y_log10()
```

```{r, fig.width = 10, fig.height=7.5}
ggplot(meta) +
  geom_jitter(aes(x = orig.ident, y = nCount_RNA, size = dilution, color = dilution), width = .2) +
  scale_y_log10()
```

```{r, fig.width = 10, fig.height=7.5}
ggplot(meta) +
  geom_col(aes(x = orig.ident, y = nFeature_RNA, fill = dilution), position = "dodge") +
  scale_y_log10()
```


## Load the unmapped read counts

```{r}
all_reads <- fread("../../data/no_reads")

all_reads[, orig.ident := gsub("[0-9].*", "", V1)]

all_reads[, dilution := gsub("[A-Z][a-z]?", "", V1)
         ][, dilution := gsub("_.*", "", dilution)
           ][, dilution := gsub("100k", "", dilution)]

all_reads[, replicate := gsub("_S[0-9].*", "", V1)
         ][, replicate := gsub(".*_", "", replicate)]

```






```{r}
subset(dil, idents = "1e7")@assays$RNA$counts
```

### Rough UMI depth
```{r}
row_sums <- data.table(rowSums(subset(dil, subset = dilution == "1e7")@assays$RNA$counts), check.names = T )

ggplot(row_sums, aes(x = "1e7", y = V1)) +
  geom_violin() +
  geom_quasirandom() +
  scale_y_log10()


```

### Origin
```{r}
meta 
```







### Remove undetermined

```{r}
# dil <- subset(dil, 
= sample != "Undetermined")
```

### Insert gene names
```{r insert gene names}
dil.id <- dil

# prepare the gene names-id table
gt <- data.table("Geneid" = rownames(dil)) # gene title

gt_s <- setDT(grch38[, c("ensgene", "symbol")])
colnames(gt_s)[1] <- "Geneid"

gt_join <- gt_s[gt, on = .(Geneid), mult = 'first'][symbol %in% NA | symbol == "", symbol := Geneid]
gt_sym <- gt_join[, symbol]
gt_id <- gt_join[, Geneid]

# insert gene names
dil.id <- dil
rownames(dil) <- gt_sym
# rm(list = c("gt", "gt_s", "gt_sym"))
```



## Initial QC


### Aggregated counts per patient



```{r, fig.height=5, fig.width=9}
meta.dil <- data.table(dil@meta.data)

## Tables
# knitr::kable(caption = "Aggregated RNA counts per patient", align = "c",
#   meta.dil[, .("RNA count" = sum(nCount_RNA)), by = .("patient" = orig.ident)][order(as.numeric(gsub("K", "", patient)))])
# 
# knitr::kable(align = "c",
#   meta.dil[, .("RNA count" = mean(nCount_RNA)), by = .("patient" = orig.ident, fraction)
#            ][order(as.numeric(gsub("K", "", patient)))])

stat <- meta.dil[, .("RNA count" = sum(nCount_RNA)), by = .("patient" = orig.ident, fraction)]
stat[patient == "Undetermined", patient := "unknwn"]

stat$patient <- factor(stat$patient, levels = unique(stat[order(as.numeric(gsub("K", "", patient))), patient]))
stat$fraction <- factor(stat$fraction, levels = c("RBC", "WBC", "P", "AB", "MV", "EX", "SP", "FT"))
stat$sample <- factor(stat$sample, levels = unique(stat[order(as.numeric(sample)), sample]))


sample_count <- meta.dil[, .(.N, "RNA count" = sum(nCount_RNA)), by = .("patient" = orig.ident)]
sample_count[patient %notlike% "Und", N := N/8]
sample_count$patient <- factor(sample_count$patient, levels = unique(sample_count[order(as.numeric(gsub("K", "", patient))), patient]))
sample_count[patient == "Undetermined", patient := "unknwn"]

ggplot(stat) +
  geom_col(aes(x = patient, `RNA count`)) +
  geom_text(data = sample_count, aes(label = N, x = patient, y = `RNA count`), vjust = -.5) +
  theme_cowplot(12) +
  ggtitle("aggregated RNA counts") +
  theme(axis.text.x = element_text(angle = -45), plot.title = element_text(hjust = .5)) +
  scale_y_continuous(limits = c(0, 8e6), expand = c(0, 0))
```


```{r, fig.height=5, fig.width=10}
ggplot(stat) +
  geom_col(aes(x = patient, `RNA count`, fill = fraction)) +
    geom_text(data = sample_count, aes(label = N, x = patient, y = `RNA count`), vjust = -.5) +
  theme_cowplot(12) +
  ggtitle("aggregated RNA counts") +
  theme(axis.text.x = element_text(angle = -45), plot.title = element_text(hjust = .5)) +
  scale_y_continuous(limits = c(0, 8e6), expand = c(0, 0))
```

```{r}
meta.dil <- meta.dil[orig.ident %notlike% "Und"]

stat <- meta.dil[, .("RNA count" = sum(nCount_RNA)), by = .("patient" = orig.ident, sample)]
stat[patient == "Undetermined", patient := "unknwn"]


stat$patient <- factor(stat$patient, levels = unique(stat[order(as.numeric(gsub("K", "", patient))), patient]))
stat$sample <- factor(stat$sample, levels = unique(stat[order(as.numeric(sample)), sample]))

ggplot(stat) +
  geom_jitter(aes(x = patient, `RNA count`, color = sample),
              size = 3, width = .1, height = 0) +
  theme_cowplot(12) +
  ggtitle("aggregated RNA counts") +
  theme( plot.title = element_text(hjust = .5))
```

```{r replicate difference}
rep1 <- meta.dil[replicate == 1, .("RNA count 1" = nCount_RNA, sample, "patient" = orig.ident, fraction)]
rep2 <- meta.dil[replicate == 2, .("RNA count 2" = nCount_RNA, sample, "patient" = orig.ident, fraction)]

rep <- rep1[rep2, on = c("sample", "patient", "fraction")][, diff := (`RNA count 1` - `RNA count 2`)]


rep$patient <- factor(rep$patient, levels = unique(rep[order(as.numeric(gsub("K", "", patient))), patient]))
rep$sample <- factor(rep$sample, levels = unique(rep[order(as.numeric(sample)), sample]))

ggplot(rep, aes(x = patient, y = diff, color = sample, shape = fraction)) +
  geom_jitter(size = 3, width = .1, height = 0) +
  scale_shape_manual(values=seq(0,15))  +
  theme_cowplot(12) +
  ggtitle("inter-replicate RNA counts difference") +
  theme(plot.title = element_text(hjust = .5)) 
```
### Aggregated gene counts per patient

```{r, fig.height=5, fig.width=9}
meta.dil <- data.table(dil@meta.data)

## Tables
# knitr::kable(caption = "Aggregated gene counts per patient", align = "c",
#   meta.dil[, .("gene count" = sum(nFeature_RNA)), by = .("patient" = orig.ident)][order(as.numeric(gsub("K", "", patient)))])
# 
# knitr::kable(align = "c",
#   meta.dil[, .("gene count" = mean(nFeature_RNA)), by = .("patient" = orig.ident, fraction)
#            ][order(as.numeric(gsub("K", "", patient)))])

stat <- meta.dil[, .("gene count" = sum(nFeature_RNA)), by = .("patient" = orig.ident, fraction)]
stat[patient == "Undetermined", patient := "unknwn"]

stat$patient <- factor(stat$patient, levels = unique(stat[order(as.numeric(gsub("K", "", patient))), patient]))
stat$fraction <- factor(stat$fraction, levels = c("RBC", "WBC", "P", "AB", "MV", "EX", "SP", "FT"))

sample_count <- meta.dil[, .(.N, "gene count" = sum(nFeature_RNA)), by = .("patient" = orig.ident)]
sample_count[patient %notlike% "Und", N := N/8]
sample_count$patient <- factor(sample_count$patient, levels = unique(sample_count[order(as.numeric(gsub("K", "", patient))), patient]))
sample_count[patient == "Undetermined", patient := "unknwn"]

ggplot(stat) +
  geom_col(aes(x = patient, `gene count`)) +
  geom_text(data = sample_count, aes(label = N, x = patient, y = `gene count`), vjust = -.5) +
  theme_cowplot(12) +
  ggtitle("aggregated gene counts") +
  theme(axis.text.x = element_text(angle = -45), plot.title = element_text(hjust = .5)) +
  scale_y_continuous(limits = c(0, 8.5e5), expand = c(0, 0))
```


```{r, fig.height=5, fig.width=10}
ggplot(stat) +
  geom_col(aes(x = patient, `gene count`, fill = fraction)) +
    geom_text(data = sample_count, aes(label = N, x = patient, y = `gene count`), vjust = -.5) +

  theme_cowplot(12) +
  ggtitle("aggregated gene counts") +
  theme(axis.text.x = element_text(angle = -45), plot.title = element_text(hjust = .5)) +
  scale_y_continuous(limits = c(0, 8.5e5), expand = c(0, 0))
```

```{r}
meta.dil <- meta.dil[orig.ident %notlike% "Und"]

stat <- meta.dil[, .("gene count" = sum(nFeature_RNA)), by = .("patient" = orig.ident, sample)]
stat[patient == "Undetermined", patient := "unknwn"]


stat$patient <- factor(stat$patient, levels = unique(stat[order(as.numeric(gsub("K", "", patient))), patient]))
stat$sample <- factor(stat$sample, levels = unique(stat[order(as.numeric(sample)), sample]))

ggplot(stat) +
  geom_jitter(aes(x = patient, `gene count`, color = sample),
              size = 3, width = .1, height = 0) +
  theme_cowplot(12) +
  ggtitle("aggregated gene counts") +
  theme( plot.title = element_text(hjust = .5))
```

```{r replicate difference}
rep1 <- meta.dil[replicate == 1, .("gene count 1" = nFeature_RNA, sample, "patient" = orig.ident, fraction)]
rep2 <- meta.dil[replicate == 2, .("gene count 2" = nFeature_RNA, sample, "patient" = orig.ident, fraction)]

rep <- rep1[rep2, on = c("sample", "patient", "fraction")][, diff := (`gene count 1` - `gene count 2`)]


rep$patient <- factor(rep$patient, levels = unique(rep[order(as.numeric(gsub("K", "", patient))), patient]))
rep$sample <- factor(rep$sample, levels = unique(rep[order(as.numeric(sample)), sample]))

ggplot(rep, aes(x = patient, y = diff, color = sample, shape = fraction)) +
  geom_jitter(size = 3, width = .1, height = 0) +
  scale_shape_manual(values=seq(0,15))  +
  theme_cowplot(12) +
  ggtitle("inter-replicate gene counts difference") +
  theme(plot.title = element_text(hjust = .5)) 
```


### Features
```{r percentage_features}
dil <- PercentageFeatureSet(dil, pattern = "^MT-", col.name = "percent_mt")
dil <- PercentageFeatureSet(dil, "^RP[SL]", col.name = "percent_ribo")
dil <- PercentageFeatureSet(dil, "^HB[^(P)]", col.name = "percent_hb")
dil <- PercentageFeatureSet(dil, "PECAM1|PF4|GP9|PPBP|ITGA2B|GP1BA", col.name = "percent_plat")
```

```{r features-fractions, fig.height= 10, fig.width=15}
VlnPlot(dil, features = c('nCount_RNA','nFeature_RNA', 'percent_mt', 'percent_hb', "percent_ribo", "percent_plat"), 
        group.by = "fraction")
```

#### Further QC

```{r}
FeatureScatter(dil, "nCount_RNA", "nFeature_RNA", pt.size = 1, plot.cor = T) + scale_x_continuous(labels = scales::scientific) + NoLegend()
```




### Preprocessing

**No preprocessing is necessary at this point. Filtering out reads from blood fractions is not established (yet).**



### Save the output

```{r save the data}

# SaveH5Seurat(dil.id, filename = "output/kidney_ev_seq_preproc.h5Seurat", overwrite = TRUE, verbose = T)
saveRDS(dil.id, file = "output/kidney_ev_seq_preproc.Rds")
```


### Session info

```{r}
sessionInfo()
```


